---
title: "rGriffin"
author: 
- name: Mariana E Martinez-Sanchez
- name: Stalin Muñoz
- name: Miguel Carrillo
- name: Eugenio Azpeitia
- name: David Rosenblueth
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rGriffin}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction


# Queries

## Create a query

## Add restrictions to the query

## Run the query



# Connect to other packages

## BoolNet

The R package includes various functions to import and export data to `BoolNet`.

It is possible to obtain the topology of a `BooleanNetwork` object using `get.net.topology()`. This function determines the sign of each regulation as positive '+', negative '-' or ambiguous 'MA'. All regulations between nodes are considered mandatory. The function can also detect Non-functional regulations 'NR'.

```
library(BoolNet)
data("cellcycle")
topology <- get.net.topology(cellcycle)
topology
```

It is also possible to convert an `AttractorInfo` object into a data.frame using `attractor2dataframe`.

```
attr <- getAttractors(cellcycle)
attr.df <- attractor2dataframe(attr)
attr.df
```

Using this functions, it is possible to create a query that corresponds to a Boolean network.

```
q <- create.gquery.graph(topology, cellcycle$genes)
q <- add.gquery.attractors(q, attr.df[cellcycle$genes])
print(q)
```

Export to BoolNet

run.gquery

dataframe2attractor


## Other

It is possible to plot the network topology dataframe with the R package igraph. This dataframe can also be used to import and export the network topology to other resources like the python library networkx or to the software cytoscape.

It is possible to export the network functions as an SBML file using the BoolNet function toSBML().





# References



Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
